The title

### 2024-08-23 manuscript.Rmd
working on intro - kind of a mess lol

[@dplyr]

@@

>>>>

Example citation: [@dplyr]

@@

Preliminary Outline

Intro - awareness of death, especially in light of COVID.

Various existential philosophers and psychologists have proposed ways in which we deal with the awareness of death and the anxiety this awareness often causes. Psychoanalyst Erik Erikson proposed that during mid-life one becomes acutely aware of their oncoming death and is motivated to care for things which will outlast themselves. He called this act of caring generativity (CITATION). In *The Denial of Death* (1973), philosopher Ernest Becker posits that humans undertake immortality projects to curb their sense of vulnerability to death. Similarly, psychiatrist Robert Jay Lifton (1979), a mentee of Erikson, described the awareness of death as being ever present and motivating us to create symbols thereby allowing ourselves to imagine us as symbolically immortalized. Existential psychiatrist, Irvin Yalom notes that many of his clients experiencing anxiety about their death take comfort in “rippling”, the idea that one’s lasting effects on the world will ripple out and influence the world after they have died (CITATION).

Although these thinkers use different terminology, there are several common themes among their ideas. (1) Our physical death is an inevitability, and we often find our awareness of its inevitability to be aversive: that is, it can cause angst, death-anxiety, despair, being-towards-death, terror, and so on. However, (2) we take comfort in the idea that other, non-physical parts of us continue to exist indefinitely after our biological death, through mechanisms such as the heroic archetype and symbolic self. (3) Finally, we can take action to promote these non-physical aspects of the self, such as through search for meaning, immortality project, care, generativity, rippling.

Sense of immortality. Originally theorized by Lifton 1979. Awareness of death drives a fundamental human need to "develop a sense of continuity and lastingness" beyond the lifespan. Some denial of death is thought to be universal, according to Lifton. But acceptance of the reality of death can drive "personal and collective creativity" (aka generativity?). Humans as meaning-seeking creatures. (The above was taken from Drolet 1990).

Lifton's five modes of symbolic immortality: (1) biological or biosocial mode (ancestors and progeny, but can be extended to one's culture, values, and social identity), (2) creative mode (one's work or social contribution, influence), (3) natural mode (being a part of the broader world transcending oneself), (4) spiritual mode (spiritual attainment, possibly a literal sense of an afterlife but not always, search for ultimate meaning), (5) mode of experiential transcendence (ability to lose oneself in what might be called a "flow" experience). The fifth mode must occur in the context of one of the other four to really be considered transcendent, but it has a great capacity to bring about personal change.

Erikson's theory and generativity.

Current measures of generativity are more about domains of generativity, actions, trait. Loyola Generativity Scale

Lacking measure of the underlying motivation, thought to be related to a motive to expand the sense of self beyond the lifetime (Kotre 1984). This is related to sense of symbolic immortality. McAdams and de St. Aubin (1992) found symbolic immortality to be one of five themes across content analysis of autobiographical episodes.

Terror Management Theory and generativity/symbolic immortality. Claims of how we suppress death-anxiety have been investigated experimentally, primarily through the paradigm of Terror Management Theory (TMT). Based on the theories of Ernest Becker, TMT posits that human awareness of death is always present to some degree. This awareness of our inevitable death, coupled with a strong aversion to thoughts of death, causes terror and is pushed out of our consciousness by our creation of meaning systems (CITATION). TMT refers to a person's awareness of death as mortality salience (MS). The MS hypothesis of TMT posits than an increase in one's awareness of death causes an increase in compensatory behaviors to lower their death anxiety, either by distracting from the awareness of death or by the promotion of meaningful cultural worldviews. A meta-analysis of 277 experiments has found mortality salience to have moderate effects (*r* = 0.35; Burke, Martens & Faucher, 2010). TMT literature has found three consistent buffers of our awareness of death: self-esteem, interpersonal relationships, and cultural worldview (CITATION, unless it's the same as above, in which case make that clear).

TMT research is conducted in two paradigms. In the MS paradigm, a participant’s increased awareness of death causes an increase in compensatory buffers. In the death-thought accessibility paradigm, experimentally decreasing participants’ buffers, such as lowering self-esteem or challenging their worldview beliefs, causes (CHECK THIS AND CITE ACTUAL STUDIES) an increase in their awareness of death. Together, these experiments provide convincing evidence of these of suppressing awareness of death and death anxiety. It is assumed that such changes in attitude and behavior suppress thoughts of death by giving a sense of existential immortality; however, to our knowledge no researchers have attempted to assess subjective feelings of immortality. This missing literature may in part be because of a lack of a viable measure of existential immortality. We view this dearth as a large gap in the literature and methodology.

Sense of psychological death?

Somewhat loss of religiosity in the West that may have been a way of coping/finding a sense of immortality in the past. A sense of collective/social immortality may be a way to address this?

Two attempts have been made to develop such a measurement. Drolet (1990) developed the Sense of Symbolic Immortality Scale. This scale was based on Robert J. Lifton’s theory of symbolic immortality in which a sense of symbolic immortality can be achieved through five different modes: the biological mode in which one lives on through their genetic progeny, the creative mode in which one’s accomplishments outlast oneself, the natural mode in which one feels they are a part of the universe, the spiritual mode in which one transcends the biological and physical realm to a higher spiritual realm to transcend death, and the mode of experiential transcendence in which one experiences a phenomenological state of flow. The Sense of Symbolic Immortality Scale contains five subscales corresponding to each proposed mode.

[Is Matthews and Kling a thing???]

Drolet studied 136 adults, 18-30 and 30-40 year olds, and hypothesized that those in their 30's (established adults) would have a greater sense of symbolic immortality than the young adults (18-30). The measure is inherently subjective, not only by the nature of self-report, but in that what we are interested in is what a person *believes* and how they *feel* about these subjects. The scale as a whole had a high internal consistency (alpha = .91) and test-retest reliability was r = .97. Internal consistency of subscales for the five theoretical modes of immortality was mixed. Of the five, spiritual immortality was the most distinct from the scale as a whole and the other subscales. (Note that this seems relevant to what we found in our own scale development. For more in the Discussion later on.) Factor analysis showed three factors, mapping onto biosocial, creative, and spiritual. The transcendent and natural items may be related to biosocial.

Moving beyond the scale development itself (still Drolet 1990), SSI correlated negatively with death anxiety (Templer's Death Anxiety Scale) and had a strong (r = .84) positive relationship with purpose in life (Maholick's Purpose in Life Test). In interpreting the very strong correlation, the author suggests that SSI is a broader construct than Purpose in Life and the scale itself may be less prone to social desirability effects than the PIL, although this had not been directly tested. Age group was also related, with established adults having a higher SSI, particularly in the biosocial and creative domains.

We see multiple issues with using the Symbolic Immortality Scale. First, the study was underpowered, conducting exploratory factor analysis of 67 items using a sample of 136. Second, the scale was developed in French, and we do not take for granted the psychometric properties of a translated version. Third and most fundamentally, the scale has poor face validity and appears to measure the constructs theorized to symbolically immortalize rather than a sense of symbolic immortality directly. For example, the scale includes items such as “My sex life contributes greatly to my well-being”, “Intimate relationships scare me”, and “I am sure of who I am." Although related to the constructs (such as interpersonal relationships and self-esteem) which theoretically help cope with death, it is unclear how these items represent the construct of symbolic immortality itself.

Mathews and Mister (1988) also developed a scale involving symbolic immortality, sensation seeking, and psychic numbness. Study of 400 adults. Experiential transcendence was operationalized as something like Zuckerman's (1979) sensation seeking, which may not fully capture the original intent (the experience of losing oneself). Items were forced into five factors, and the five factors largely aligned with Lifton's constructs. Although internal consistency was at least acceptable for each factor, goodness of fit statistics are not reported. The items themselves seem to have more face validity than the scale by Drolet, but some factors seem more behavioral and specific: pertaining to one's religiosity or biological children. The Nature and Creative factors seems most useful and theoretically aligned with Lifton.

The goal of the present research is to develop a scale that more directly measures one’s sense of existential immortality using a larger sample and current best practices for scale development.
